---
title: "tidytuesday_2025_10_07"
output:
  html_document: default
  pdf_document: default
date: "2025-10-07"
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EuroLeague Basketball 

This week’s #TidyTuesday dataset explores EuroLeague Basketball—the top-tier men’s club competition in Europe, where national champions and powerhouses meet on one court.
I’m focusing on arenas—the question that captures the scale of the game:
 "Which countries boast the biggest EuroLeague arenas?"
Each team’s home court tells a story of local passion and infrastructure: from 3,000-seat halls in historic cities to 20,000-seat mega-arenas in basketball-mad capitals. We can compare how seating size varies across nations, explore whether larger venues align with winning records, and even visualize how geography shapes basketball fandom.
Thanks to Natasa Anastasiadou for curating this week’s dataset!

# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load( tidyverse, tidytuesdayR, DataExplorer, tools, scales, janitor,ggbump,patchwork, slider) 

```

# Import the Data
```{r load_data, echo=FALSE}

tuesdata <- tidytuesdayR::tt_load('2025-10-07')

euroleague_basketball <- tuesdata$euroleague_basketball


# Explore structure and completeness
introduce(euroleague_basketball  )


glimpse(euroleague_basketball )



```

# Data Prep
```{r}
dat <- euroleague_basketball %>%
  clean_names() %>%
  mutate(
    across(c(team, country, arena), ~ stringr::str_squish(.x))
  )

# Parse capacities
cap_long <- dat %>%
  mutate(cap_tokens = stringr::str_extract_all(capacity, "\\d[\\d,\\.]*")) %>%
  tidyr::unnest_longer(cap_tokens, values_to = "cap_token") %>%
  mutate(capacity_num = readr::parse_number(cap_token)) %>%
  filter(!is.na(capacity_num))

# Max per team (if multiple arenas), then max per country
country_max <- cap_long %>%
  group_by(country, team) %>%
  summarise(team_max_capacity = max(capacity_num), .groups = "drop") %>%
  group_by(country) %>%
  summarise(max_capacity = max(team_max_capacity), .groups = "drop") %>%
  arrange(desc(max_capacity)) %>%
  mutate(
    max_capacity_lab = label_comma()(max_capacity)
  )

# Lollipop plot 
ggplot(country_max,
       aes(x = max_capacity, y = forcats::fct_reorder(country, max_capacity))) +
  geom_segment(aes(x = 0, xend = max_capacity, y = country, yend = country),
               linewidth = 1, alpha = 0.6) +
  geom_point(size = 3) +
  geom_text(aes(label = max_capacity_lab), hjust = -0.15, size = 3.5) +
  scale_x_continuous(
    labels = label_comma(),                 # stable across {scales} versions
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Maximum EuroLeague Arena Capacity by Country",
    subtitle = "Each Country’s Largest Listed Home-Arena Capacity",
    x = "Max Capacity (Seats)", y = NULL,
    caption = "Source: EuroleagueBasketball"
  ) +
 theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "gold", colour = NA),
    plot.background  = element_rect(fill = "gold", colour = NA),
    axis.text.y  = element_text(size = 12, face = "bold", family = "Arial",colour = "black"),
    axis.title.y = element_text(size = 12, face = "bold", family = "Arial"),
    axis.title.x = element_text(size = 12, face = "bold", family = "Arial"),
    strip.text   = element_text(face = "bold", size = 13),
    plot.title   = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle= element_text(size = 12, face = "bold", hjust = 0.5, lineheight = 1.2),
    axis.text.x  = element_text(size = 8, face = "bold", family = "Arial", hjust = 1)
  )


# Optional:
 ggsave("country_max_capacity_lollipop.png", width = 9.5, height = 7.5, dpi = 320)
 
 

```


# Key Insights
- Neck-and-neck at the top: Serbia (18,386) edges Greece (18,300) by just 86 seats—effectively a tie for largest EuroLeague arena capacity.
- Clear upper tier: United Arab Emirates (17,000) rounds out the top three, signaling strong venue investment in a newer market.
- Core capacity band: Most countries cluster between ~11,500 and 15,700 seats (France, Spain, Lithuania, Turkey, Italy, Israel, Germany)—a tight mid-range.
- Takeaway: If you’re chasing the loudest, largest crowds, look to Belgrade and Athens—this week’s tidytuesday snapshot suggests they set the bar for EuroLeague scale.
