---
title: "tidytuesday_2025_09_09"
output: html_document
date: "2025-09-09"
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Henley Passport Index: Focus on Africa
This week I’m exploring global mobility through the Henley Passport Index with a specific focus on African passports. The Index counts how many destinations each passport can enter without pre-departure government approval. For every origin–destination pair, a score of 1 is assigned if travelers can enter visa-free, with a visa on arrival, a visitor’s permit, or an electronic travel authority (ETA); a score of 0 applies when an e-Visa (or any pre-approved visa) is required. A passport’s total score is the sum of destinations with value 1. Rankings are updated monthly and reflect global ranks—recent shifts have drawn notable attention.
Question to explor
Which African countries posted the biggest gains (or losses) in visa-free access?
Thanks to Brenden Smith and Jen Richmond for curating this week’s dataset.

# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load( tidyverse, tidytuesdayR, DataExplorer, tools, scales, janitor,ggbump,patchwork) 

```



# Import the Data
```{r load_data, echo=FALSE}

tuesdata <- tidytuesdayR::tt_load('2025-09-09')

country_lists <- tuesdata$country_lists
rank_by_year <- tuesdata$rank_by_year

# Explore structure and completeness
introduce(country_lists )
introduce(rank_by_year)

glimpse(country_lists )
glimpse(rank_by_year )


```



# Data Prep


```{r data_prep, echo= FALSE}

africa_data <- rank_by_year %>%
  filter(region == "AFRICA") %>% 
  drop_na()

top_3 <- africa_data %>%
  filter(year == 2025) %>%
  slice_min(order_by = rank, n = 3, with_ties = TRUE) %>%
  arrange(rank)

pad <- 0.6


```




# Plot 
## Bump plot
```{r bump_plot, echo=FALSE}

passport_bump_plot <-
  ggplot(africa_data,
         aes(x = year, y = rank, group = country, alpha = -rank, colour = -rank)) +
  geom_bump(size = 1,  show.legend = FALSE) +
  geom_point(size = 2, shape = 21, fill = "white", stroke = 1.2, show.legend = FALSE) +
  scale_color_gradientn(
    colours = c("#deebf7", "#00008B", "#000000"),
    values  = scales::rescale(c(0, 0.8, 1)),
    guide   = "none"
  ) +
  geom_text_repel(
    data = top_3,
    aes(label = country, colour = -rank),
    nudge_x = 0.3, hjust = 0, direction = "y",
    min.segment.length = Inf, seed = 42, size = 3.8,
    show.legend = FALSE
  ) +
  scale_y_reverse(limits = c(120, 1), breaks = seq(1, 120, by = 9)) +
  scale_x_continuous(
    breaks = c(2006, 2010, 2015, 2020, 2025),
    expand = expansion(mult = c(0.02, 0.14))
  ) +
  coord_cartesian(xlim = c(2006, 2025 + pad), clip = "off") +
  labs(  title = "African Passport Power (2006–2025): Seychelles & Mauritius lead; \n South Africa follows. Most never hit the global top 50. ",
  subtitle = "Seychelles edges upward to new highs, Mauritius dips mid-2010s then recovers, \n and South Africa breaks a long plateau with a 2025 upswing.",
  caption = "Data Source: https://www.henleyglobal.com/passport-index/", 
  x = "Year", y = "Global Rank (Lower is Better)") +
  

  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "yellow", colour = NA),
    plot.background  = element_rect(fill = "yellow", colour = NA),
    axis.text.y  = element_text(size = 8, face = "bold", family = "Arial"),
    axis.title.y = element_text(size = 12, face = "bold", family = "Arial"),
    axis.title.x = element_text(size = 12, face = "bold", family = "Arial"),
    strip.text   = element_text(face = "bold", size = 13),
    plot.title   = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle= element_text(size = 12, face = "bold", hjust = 0.5, lineheight = 1.2),
    axis.text.x  = element_text(size = 8, face = "bold", family = "Arial", angle = 45, hjust = 1)
  )

passport_bump_plot


```

```{r save_plot, echo=FALSE}
ggsave(
  plot =passport_bump_plot ,
  filename = "passport_bump_plot.png",
  width = 12,    
  height = 6.3,  
  units = "in",
  dpi = 600
)

```




# Key Insights

- Seychelles ( 156 destinations) and Mauritius (149) dominate Africa’s passport power.
- South Africa rebounds post-COVID (103), breaking a long plateau.
- Most countries cluster around r50–90, with gradual gains, not leaps.
- COVID effect: Clear dip in 2020–21, followed by recovery into 2023–25.
 -Access gap: Top passports offer ~150+ destinations—far above the continental median.















`










```{r}
#sessioninfo()
```


