---
title: "tidytuesday_2025_09_30"
output:
  html_document: default
  pdf_document: default
date: "2025-09-30"
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cranes in Sweden: Timing Is Everything
This week’s #TidyTuesday dataset comes from long-term crane counts at Lake Hornborgasjön in Västergötland, Sweden. For over 30 years, dedicated counters have tracked the spectacular spring migration, recording when tens of thousands of cranes stop at the lake on their way north.
I’m focusing on timing—the question birders and visitors always ask:
🕊️ “When is the best time of year to see thousands of cranes?”
With daily counts from 1994–2024, we can map out spring arrivals, identify peak weeks, and see how weather patterns shift the season earlier or later. Given the literature on Crane migration, I expected the magic window to fall in late March to early April
Thanks to Carl Borstell for curating this week’s dataset.

# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load( tidyverse, tidytuesdayR, DataExplorer, tools, scales, janitor,ggbump,patchwork, slider) 

```


# Import the Data
```{r load_data, echo=FALSE}

tuesdata <- tidytuesdayR::tt_load('2025-09-30')

cranes<- tuesdata$cranes


# Explore structure and completeness
introduce(cranes )


glimpse(cranes)




```

# Data Prep
```{r}
cranes_clean <- cranes %>%
  mutate(Date = lubridate::ymd(date)) %>%
  # I filtered for the spring migration 
  filter(month(Date) %in% c( 3, 4, 5)) %>%
  # Calculate Day of Year (DOY) to normalize across years
  mutate(
    DOY = yday(Date),
    Year = year(Date),
    Day_of_Spring = as.numeric(Date - ymd(paste0(Year, "-03-01")))
  ) %>%
  select(Year, Date, observations, DOY, Day_of_Spring) %>%
  filter(observations > 0)
```



# Plot

```{r}

# Spring heatmap (year × day-of-year)
crane_peak_plt <- cranes_clean |>
  ggplot(aes(DOY, factor(Year), fill = observations)) +
  geom_tile() +
  scale_fill_viridis_c(labels = c("0", "5,000", "10,000", "15,000", "20,000", "25,000+"),
                       option = "plasma", 
                        breaks = c(0, 5000, 10000, 15000, 20000, 25000),
    ) +
  labs(
    title = "The Two-Week Window: Charting 30 Years of Peak Crane Migration",
subtitle=	"Daily observations confirm the highest crane populations consistently \n arrive between April 1st and April 15th at Lake Hornborgasjön",
    x = "Day of Year", y = "Year", fill = "Count",  

  ) + scale_x_continuous(limits = c(60, 135), breaks = seq(60, 135, by = 15),
                         labels = c("Mar 01", "Mar 15", "Apr 01", "Apr 15", "May 01", "May 15") ) +
  theme_minimal(base_size = 12) +
theme(
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "lightgray", colour = NA),
   plot.background  = element_rect(fill = "lightgray", colour = NA),
    axis.text.y  = element_text(colour= "black", size = 8, face = "bold", family = "Arial"),
    axis.title.y = element_text(size = 12, face = "bold", family = "Arial"),
    axis.title.x = element_text(size = 12, face = "bold", family = "Arial"),
    strip.text   = element_text(face = "bold", size = 13),
    plot.title   = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle= element_text(size = 12, face = "bold", hjust = 0.5, lineheight = 1.2),
    axis.text.x  = element_text(size = 12, face = "bold", family = "Arial",  hjust = 1)
  )

crane_peak_plt
```


# Save plot
```{r save_plot, echo=FALSE}
ggsave(
  plot =crane_peak_plt ,
  filename = "crane_peak_plt.png",
  width = 12,    
  height = 6.3,  
  units = "in",
  dpi = 600
)

```

# Key Insights

- Consistent peak window: Across three decades, the densest crane counts cluster tightly between April 1st and April 15th.
- High concentration: This two-week span reliably produces the highest daily counts (often exceeding 20,000–25,000 cranes).
- Year-to-year variation: The exact peak shifts a little earlier or later depending on spring conditions, but the central window is remarkably stable.
- So, if your goal is to see tens of thousands of cranes, plan a visit in early April—that’s the sweet spot for witnessing the migration spectacle.

