---
title: "tidytuesday_2025_07_01"
output: html_document
date: "2025-07-01"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Weekly US Gas Prices

This week we're exploring weekly US gas prices! The data comes from the U.S. Energy Information Administration (EIA), which publishes average retail gasoline and diesel prices each Monday. The original data (including additional datasets) can be found at eia.gov/petroleum/gasdiesel, and the weekly time series used here was downloaded from this XLS file.

Gas price methodology:

"Every Monday, retail prices for all three grades of gasoline are collected mainly by telephone and email from a sample of approximately 1,000 retail gasoline outlets. The prices are published around 5:00 p.m. ET Monday, except on government holidays, when the data are released on Tuesday (but still represent Monday's price). The reported price includes all taxes and is the cash pump price paid by a consumer as of 8:00 a.m. Monday. This price represents the self-serve price except in areas having only full-serve. The price data from the sample are used to calculate volume-weighted average gasoline price estimates at the national, regional, and selected city and state levels for all gasoline grades and formulations."
Diesel price methodology:

"Every Monday, cash self-serve on-highway diesel prices (including taxes) are collected from a sample of approximately 590 retail diesel outlets in the continental United States. The sample includes a combination of truck stops and service stations that sell on-highway diesel fuel. The data represent the price of ultra low sulfur diesel (ULSD), which contains less than 15 parts-per-million sulfur. All collected prices are subjected to automated edit checks during data collection and data processing. Data flagged by the edits are verified with the respondents. Imputation is used for companies that cannot be contacted and for reported prices that are extreme outliers. The average survey response rate for 2020 was 98%. Average national and regional prices are released around 5:00 p.m. ET on Mondays, except on government holidays, in which case the data are released on Tuesday (but still represent Monday's price)."

How did gas prices behave during major events like the 2008 recession or COVID-19 pandemic?
Are diesel prices more or less volatile than gasoline prices?
Do different grades or formulations of gasoline follow similar trends?
Thank you to Jon Harmon, Data Science Learning Community for curating this week's dataset.





# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load(
  tidyverse, tidytuesdayR, scales, forecast, DataExplorer,
  hrbrthemes, Metrics, gganimate, gifski, lubridate
)
```

# Import Weekly Gasoline Price Data
```{r load_data}
tuesdata <- tt_load('2025-07-01')
weekly_gas_prices <- tuesdata$weekly_gas_prices
```

# Data Wrangling and Filtering
```{r wrangle_data}
weekly_gas_prices <- weekly_gas_prices %>%
  mutate(date = as.Date(date))

gas_data <- weekly_gas_prices %>%
  filter(fuel == "gasoline", grade == "regular", formulation == "all") %>%
  arrange(date) %>%
  select(date, price)
```

# Data Structure and Missingness Overview
```{r explore_structure}
introduce(weekly_gas_prices)
plot_intro(weekly_gas_prices)
plot_missing(weekly_gas_prices)
plot_str(weekly_gas_prices)
plot_bar(weekly_gas_prices, by = "fuel")
```

# Visualize Prices During Major Economic Events
```{r event_trends}
# 2008 Recession
gas_2008 <- weekly_gas_prices %>% filter(date >= "2007-01-01", date <= "2009-12-31")

# COVID-19 Pandemic
gas_covid <- weekly_gas_prices %>% filter(date >= "2019-01-01", date <= "2021-12-31")

# Plot 2008
ggplot(gas_2008, aes(date, price)) +
  geom_line(color = "blue") +
  labs(title = "Gas Prices During the 2008 Recession", x = "Date", y = "USD")

# Plot COVID-19
ggplot(gas_covid, aes(date, price)) +
  geom_line(color = "red") +
  labs(title = "Gas Prices During the COVID-19 Pandemic", x = "Date", y = "USD")

# Overall Trend
ggplot(gas_data, aes(date, price)) +
  geom_line(color = "purple") +
  labs(title = "Weekly Gasoline Prices (Regular)", x = "Date", y = "USD")
```

# Seasonal Decomposition
```{r seasonality_check}
gas_ts <- ts(gas_data$price, start = c(1990, 34), frequency = 52)

gas_data_seasonal <- gas_data %>%
  mutate(
    seasonal = decompose(gas_ts)$seasonal,
    month = month(date, label = TRUE, abbr = TRUE)
  )

gas_data_seasonal %>%
  group_by(month) %>%
  summarise(mean_seasonal = mean(seasonal, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = mean_seasonal)) +
  geom_col(fill = "#B0A8B9") +
  theme_ipsum(grid = "Y") +
  labs(title = "Seasonality in Monthly Gasoline Prices", x = "Month", y = "Seasonal Effect")
```

# Train-Test Split for Forecasting
```{r split_data}
h <- 26
gas_test <- tail(gas_ts, h)
gas_train <- head(gas_ts, -h)
```

# Model Fitting
```{r model_fits}
# ETS
model_ets <- ets(gas_train, model = "ZZZ")
fc_ets <- forecast(model_ets, h = h)

# SARIMA
model_sarima <- auto.arima(gas_train, seasonal = TRUE)
fc_sarima <- forecast(model_sarima, h = h)

# NNAR
model_nnar <- nnetar(gas_train)
fc_nnar <- forecast(model_nnar, h = h)

# Ensemble
ensemble_preds <- (fc_ets$mean + fc_sarima$mean) / 2
```

# Model Evaluation
## SARIM performed better than all models.
SARIMA consistently performs well, with lower or comparable RMSE and MAPE than SARIMA.


```{r evaluate_models}
actuals <- as.numeric(gas_test) # get actuals first

fc_results <- bind_rows(
  tibble(model = "ETS", RMSE = rmse(actuals, fc_ets$mean), MAE = mae(actuals, fc_ets$mean), MAPE = mape(actuals, fc_ets$mean) * 100),
  tibble(model = "SARIMA", RMSE = rmse(actuals, fc_sarima$mean), MAE = mae(actuals, fc_sarima$mean), MAPE = mape(actuals, fc_sarima$mean) * 100),
  tibble(model = "NNAR", RMSE = rmse(actuals, fc_nnar$mean), MAE = mae(actuals, fc_nnar$mean), MAPE = mape(actuals, fc_nnar$mean) * 100),
  tibble(model = "Ensemble (ETS + SARIMA)", RMSE = rmse(actuals, ensemble_preds), MAE = mae(actuals, ensemble_preds), MAPE = mape(actuals, ensemble_preds) * 100)
)

fc_results
```

# Plot Forecasts vs Actuals
```{r plot_forecasts}
tibble(
  date = time(gas_test),
  Actual = actuals,
  ETS = fc_ets$mean,
  SARIMA = fc_sarima$mean,
  NNAR = fc_nnar$mean,
  Ensemble = ensemble_preds
) %>%
  pivot_longer(-date, names_to = "Model", values_to = "Value") %>%
  ggplot(aes(x = date, y = Value, color = Model)) +
  geom_line(size = 1.1) +
  labs(title = "Forecast Comparison", x = "Date", y = "Price per Gallon (USD)") +
  theme_minimal()
```

# Animate Actual and Forecased Price
```{r animation_actual_forecasted, message=FALSE, warning=FALSE}
# Obtain forcast data col "Type" added
forecast_data <- tibble(
  date = seq(max(gas_data$date) + weeks(1), by = "1 week", length.out = h),
  price = as.numeric(fc_sarima$mean),
  lower = fc_sarima$lower[, 2],  # 95% CI
  upper = fc_sarima$upper[, 2],
  Type = "Forecast"
)

# The actual data frame with col "Type" added
gas_actual <- gas_data %>%
  mutate(lower = NA, upper = NA, Type = "Actual")

# Bind for animation
plot_data <- bind_rows(gas_actual, forecast_data) %>%
  arrange(date) %>%
  mutate(frame = row_number())


# I wanted to plot with confidence ribbon to show 95% CI
gas_plot <- ggplot(plot_data, aes(x = date, y = price, color = Type)) +
  geom_ribbon(
    data = filter(plot_data, Type == "Forecast"),
    aes(x = date, ymin = lower, ymax = upper),
    fill = "green", alpha = 0.2, inherit.aes = FALSE
  ) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(plot_data, Type == "Actual"),
             aes(y = price), color = "blue", size = 0.8) +
  geom_point(data = filter(plot_data, Type == "Actual"), size = 0.6, color = "blue") +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "darkgreen"))  +

  labs(
    title = "U.S. Weekly Regular Gasoline Prices with 6-Month Forecast Through December 2025",
    subtitle = "Date: {frame_along}",
    x = "Date", y = "Price per Gallon (USD)", 
    caption = "Data Source: U.S. Energy Information Administration (EIA)"
    
  ) +
  theme_minimal(base_size = 14) +
   theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0, size = 10, face = "italic")
  ) +
  transition_reveal(along = date) + view_follow(fixed_y = c(0, 6))
  
# Save the animation, I wanted viwers to focus on the forecast so I added end pause 40
anim_save("gasoline_forecast_ets_final.gif",
          animation = animate(gas_plot, duration = 30, end_pause = 40, fps = 20, width = 1080, height = 1080,nframes = 250, renderer = gifski_renderer()))

write.csv(forecast_data, file="forecast_data.csv")
```
