---
title: "tidytuesday_2025_06_17"
output: html_document
date: "2025-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# API Specs

This week we're exploring Web APIs! The lead volunteer for TidyTuesday, Jon Harmon, is developing a book and a suite of R packages around Web API usage. To support this work, Jon curated a dataset from [APIs.guru](https://apis.guru), a project that aims to build a machine-readable Wikipedia for Web APIs using the OpenAPI Specification.
Thank you to Jon Harmon and the Data Science Learning Community for curating this week's dataset.


While there are many interesting angles to explore in this data, I chose to focus on a question posed by the Posit team:

Which providers offer the most unique APIs according to APIs.guru — and what categories do those APIs fall into?

This question uncovers not just the size of each provider’s API portfolio, but also the functional diversity of their offerings across domains like cloud, media, telecom, and beyond




# Load Libraries
```{r load_libraries, echo=FALSE, include=FALSE}
library(pacman)
pacman::p_load(tidyverse, tidytext, janitor, tidytuesdayR, scales)
```

# Load Data
```{r load_data , echo=FALSE, include=FALSE}

tuesdata <- tidytuesdayR::tt_load('2025-06-17')

api_categories <- tuesdata$api_categories
api_info <- tuesdata$api_info
api_logos <- tuesdata$api_logos
api_origins <- tuesdata$api_origins
apisguru_apis <- tuesdata$apisguru_apis
```

# Initial Glimpse
```{r glimpse_data , echo=FALSE, include=FALSE}
# Preview the structure of two main datasets.
head(api_categories)
head(api_info)
names(api_info)
names(api_categories)
```

# Compare Dataset Dimensions
```{r compare_dimensions , echo=FALSE, include=FALSE}
# Comparing the number of rows and columns in api_info and api_categories to assess alignment before joining

dim_comparison <- data.frame(
  DataFrame = c("api_info", "api_categories"),
  Rows = c(nrow(api_info), nrow(api_categories)),
  Columns = c(ncol(api_info), ncol(api_categories))
)
dim_comparison
```

# Identify Unmatched API Names
```{r unmatched_names , echo=FALSE, include=FALSE}
# Explore mismatched 'name' entries to understand missing joins between datasets

missing_categories <- anti_join(api_info, api_categories, by = "name")
missing_info <- anti_join(api_categories, api_info, by = "name")

nrow(missing_categories) # 188 
head(missing_categories)
nrow(missing_info) # none  to show
head(missing_info) 
```

# Unique Provider Count & Top 10% Contribution
```{r provider_distribution,  echo=FALSE, include=FALSE}
# Quantify the number of unique providers and the API contribution of the top 10% using api_info data

api_info %>%
  summarise(unique_providers = n_distinct(provider_name)) # 676 unique providers, nrow(api_info) = 2529

provider_counts <- api_info %>%
  count(provider_name, name = "num_apis") %>%
  arrange(desc(num_apis))

top_10_percent <- ceiling(nrow(provider_counts) * 0.10)

top_10_providers <- provider_counts %>%
  slice_head(n = top_10_percent)

total_apis <- sum(provider_counts$num_apis)
top_10_apis <- sum(top_10_providers$num_apis)
share_top_10_providers <- (top_10_apis / total_apis) * 100

cat("Top 10% providers contribute", round(share_top_10_providers, 2), "% of all APIs.\n")
```



# Top 10 Providers by API Count after Joining with Category Data 
```{r top_10_providers, echo=FALSE, include=FALSE}
# Identify the top 10 providers by total APIs offered

top_10_providers <- api_info %>%
  count(provider_name, name = "num_apis") %>%
  arrange(desc(num_apis)) %>%
  slice_max(n = 10, order_by =num_apis, with_ties=FALSE) #  Two providers tied with the 10th in API count but were excluded to maintain a clear top 10

total_apis <- api_info %>% nrow()
top_10_total <- sum(top_10_providers$num_apis)
top_10_api_share <- top_10_total / total_apis * 100

cat("Top 10 providers contribute", round(top_10_api_share, 2), "% of all APIs.\n")
```




# Join and Analyze by Category
```{r join_category_data , echo=FALSE, include=FALSE}
# Merge category data and analyze API counts by provider and category

api_info_cat <- left_join(api_info, api_categories, by = "name")

duplicates <- api_info_cat %>%
  count(name) %>%
  filter(n > 1) %>%
  arrange(desc(n))

head(duplicates)
dim(duplicates)
table(duplicates$n)

# Top 10 providers by unique API count

top_10_providers2 <- api_info_cat %>%
  group_by(provider_name) %>%
  summarise(total_apis = n_distinct(name), .groups = "drop") %>%
  slice_max(total_apis, n = 10,  with_ties = FALSE)

# Count APIs by category for these top providers

top_10_provider_cats <- api_info_cat %>%
  filter(provider_name %in% top_10_providers2$provider_name) %>%
  group_by(provider_name, apisguru_category) %>%
  summarise(num_apis = n_distinct(name), .groups = "drop") %>%
  arrange(desc(num_apis))
```

# Format Category Labels
```{r format_labels, echo=FALSE, include=FALSE}
# Clean and format provider/category names for presentation
top_10_provider_cats <- top_10_provider_cats %>%
  mutate(
    provider_name = str_to_title(provider_name),
    apisguru_category = case_when(
      is.na(apisguru_category) ~ NA_character_,
      TRUE ~ str_to_title(str_replace_all(apisguru_category, "_", " "))
    )
  )
```

# Plot: Stacked Bar by Provider and Category
```{r plot_stacked_bar , echo=FALSE,}

ggplot(top_10_provider_cats, aes(x = reorder(provider_name, -num_apis), y = num_apis, fill = apisguru_category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "API Count by Category for Top 10 Providers",
    x = "Provider",
    y = "Number of APIs",
    fill = "API Category"
  ) +
  theme_minimal(base_family = "mono") +.   # Experimenting with font family
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.title = element_text(face = "bold")
  )
```

# Final Formatting and Plot
```{r plot_stacked_bar_final, echo=FALSE}

# Ensure clean category names and reorder for plot. The x-axis labels link to the api providers

top_10_provider_cats <- top_10_provider_cats %>%
  mutate(
    apisguru_category = ifelse(apisguru_category == "Na", NA, apisguru_category),
    provider_name = str_to_title(provider_name),
    apisguru_category = case_when(
      is.na(apisguru_category) ~ NA_character_,
      TRUE ~ str_to_title(str_replace_all(apisguru_category, "_", " "))
    )
  ) %>%
  group_by(provider_name) %>%
  mutate(total_apis = sum(num_apis)) %>%
  ungroup() %>%
  mutate(provider_name = fct_reorder(provider_name, total_apis, .desc = TRUE)
  )

# Render final plot

final_plot <- ggplot(top_10_provider_cats, aes(x = provider_name, y = num_apis, fill = apisguru_category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Most of the Web’s APIs Come from Just 10 Providers ",
     subtitle = "Top 10 providers account for 62% of all APIs listed by APIs.guru (2025)",
    x = "Provider",
    y = "Number of APIs",
    fill = "API Category"
  ) +
    scale_y_continuous(breaks = seq(0, max(top_10_provider_cats$num_apis), by = 100)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, face = "bold.italic", hjust = 0.5),
    legend.title = element_text(face = "bold")
  )


ggsave("Top_10_apis_by_category.png", final_plot, width = 10, height = 6, bg="white")
```




# Summary

This week's analysis of API specifications from APIs.guru revealed a highly skewed distribution in API contributions. A small number of providers—most notably Google, AWS, Azure, and Twilio—account for the majority of listed APIs. Categories spanned cloud services, messaging, payments, and government open data. A small percentage of APIs lacked category metadata,perhaps signaling potential data quality issues. 
