---
title: "tidytuesday_2025_08_19"
output: html_document
date: "2025-08-19"
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Scottish Munros

This week’s #TidyTuesday explores Scotland’s Munros using data from the Database of British and Irish Hills (DoBIH v18.2), curated by Nicola Rennie. A Munro is a Scottish mountain over 3,000 feet; a Munro Top is a subsidiary summit also exceeding 3,000 ft but not considered a distinct mountain. The first tables were compiled by Sir Hugh Munro in 1891. Unlike other Scottish lists that require a fixed prominence (e.g., 500 ft), the Munros lack a rigid criterion—so improved surveying and editorial judgment have led to periodic reclassifications (e.g., Munro ↔ Munro Top, and occasional new additions).

In this analysis I focus on two guiding questions:
How have Munro classifications changed across editions (1891 → 1921 → 1981 → 1997 → 2021)?
Which year saw the largest number of reclassifications, and in what directions (Munro → Top vs Top → Munro)?

Data source: The Database of British and Irish Hills v18.2 (Creative Commons Attribution 4.0 International Licence). 

Thanks to Nicola Rennie for curating this week’s dataset.

This R Markdown will:
Load and tidy the TidyTuesday Munros dataset
Reshape classifications by edition and detect reclassifications:Munro ↔ Munro Top

Summarize and visualize reclassification counts by edition year
Provide fully reproducible code using tidyverse. Session info included for transparency.


# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load( tidyverse, tidytuesdayR, DataExplorer, tools, scales, janitor,openxlsx)

```


# Import the Data
```{r load_data, echo=FALSE}
# Import the #TidyTuesday data for 2025-08-19

tuesdata <- tidytuesdayR::tt_load('2025-08-19')
scottish_munros <- tuesdata$scottish_munros


# Explore structure and completeness
introduce(scottish_munros)

glimpse(scottish_munros)


```

# Prepare plot data

```{r munro_plot_data_prep, echo=FALSE}

year_cols <- names(scottish_munros) %>%
  stringr::str_subset("^(18|19|20)\\d{2}$") %>%
  as.integer() %>%
  sort() %>%
  as.character()


munros_long <- scottish_munros %>%
  select(DoBIH_number, Name, all_of(year_cols)) %>%
  pivot_longer(all_of(year_cols), names_to = "Year", values_to = "Class") %>%
  mutate(Year = as.integer(Year)) %>%
  arrange(DoBIH_number, Year)

changes_by_year <- munros_long %>%
  group_by(DoBIH_number) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(changed = Class != lag(Class)) %>%
  ungroup()

reclass_counts <- changes_by_year %>%
  filter(!is.na(changed)) %>%
  group_by(Year) %>%
  summarise(n_changes = sum(changed), .groups = "drop") %>%
  arrange(n_changes)


```


```{r, munro_plot, echo=FALSE}

munro_lollipop <- reclass_counts %>%
  mutate(Year = fct_reorder(factor(Year), n_changes, .desc = FALSE)) %>%
  ggplot(aes(x = Year, y = n_changes)) +
  geom_segment(aes(xend = Year, y = 0, yend = n_changes),
               color = "purple", linewidth = 3) +
  geom_point(size = 10, color = "purple") +
  geom_text(aes(label = n_changes), color = "white", size = 4, vjust = 0.5) +
  labs(
    title = "Munro Reclassifications by Year (1891–2021)",
    subtitle = "The biggest shake-up was in 1921 with 33 reclassifications. \n Later years (1981, 1997) saw smaller bumps, while most remained unchanged.",
    x = "Edition Year", y = "Number of Munros Reclassified",
        caption = "Source:The Database of British and Irish Hills \n
    https://www.hills-database.co.uk \n 
    Reclassification is defined as a change in status between Munro (MUN) and Munro Top (TOP) \n 
    in successive editions."

  ) +
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x =  element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    panel.grid.major.y = element_blank(),
    legend.position = "none",
   plot.caption = element_text(lineheight = 0.5),
    panel.background = element_rect(fill = "yellow", color = NA),
    plot.background  = element_rect(fill = "yellow", color = NA)
  ) +
  coord_flip()

munro_lollipop

```




```{r save_plot, echo=FALSE}
ggsave(
  plot = munro_lollipop,
  filename = "munro_lollipop.png",
  width = 12,    
  height = 6.3,  
  units = "in",
  dpi = 600
)

```


# Key Insights

- Major Shake-Up in 1921 – A total of 33 reclassifications, the largest in Munro history, reflecting Sir Hugh Munro’s substantial revisions.
- Notable Adjustments in 1981 – 16 changes occurred, marking the second-largest revision based on more accurate mapping.
- Moderate Revisions in 1997 – 10 changes under Alan Dawson’s tables, bringing the list closer to its modern form.
- Minimal Changes Elsewhere – Other years saw very few adjustments (1–3), showing overall stability in the classification.
- Recent Stability – By 2021, only one Munro was reclassified, highlighting that the list has largely settled.



```{r}
sessioninfo()
```


