---
title: "tidytuesday_2025_07_07"
output: html_document
date: "2025-07-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#The xkcd Color Survey Results

In 2010, the xkcd Color Survey asked hundreds of thousands of people to name colors they saw, revealing the different ways in which people perceive and label colors.

Color is a really fascinating topic, especially since we’re taught so many different and often contradictory ideas about rainbows, different primary colors, and frequencies of light.

Which types of users were most accurate in naming colors?
Which colors are mentioned most in the top 100 ranked color names?
Which types of users are least likely to be spam users?
Thank you to Nicola Rennie for curating this week's dataset.




# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load(
  tidyverse, tidytuesdayR, scales,, DataExplorer, gganimate, gifski, effectsize
)
```

# Import Data
```{r load_data}
tuesdata <- tt_load('2025-07-08')
users <- tuesdata$users
answers <- tuesdata$answers
color_ranks <- tuesdata$color_ranks


```

# Explore data 

## Users data
The 'users' dataset consists of 152,401 users. 
Monitor type: most users reported using LCDs. 
Y chromosome presence was reported by 71% of respondents suggesting a majority male (chromosomal) sample.
Colorblindness was reported by only 4% of users, I think this is consistent with expected population rates.
The spam probability variable has a mean of 0.22, but a maximum value of 8.43.  Since probabilities must lie within [0, 1]. Thus,  
I treated spam_prob > 0.5 as an indicator of unreliable or suspicious user data and excluded those rows from further analysis (see next section).

```{r describe_users_data, echo=FALSE}
describe(users)
```

## answers data 
Total Responses: 1,058,211
Id Range: 1 to 152,401 users indicating number of respondents 
Color Hex Codes (hex): Stored as integers internally, but represent hexadecimal color values.
Rank: Reflects how well a user matched their response to the top 5 ranked names for a given color. Range: 1 (most accurate) to 5 (least aligned with consensus)

```{r describe_answers_data, echo=FALSE}
describe(answers)
```
## color_ranks
This dataset contains the 949 most commonly named colors from the xkcd color survey.
Total Colors: 949, color, rank, and hex
Rank: Range: 1 (most frequently used color name) to 949 (least frequent).

```{r describe_ranks_data, echo=FALSE}
describe(color_ranks)
```


# Data Cleaning 

## Data Integration
I  merged user response data with demographic information using user_id.

##Missing Data Removal
I filtered out all rows with missing values for either y_chromosome or colorblind

#Spam Filtering
 I excluded users with a spam_prob greater than 0.5. This threshold was used to remove respondents likely to have submitted unreliable or non-serious data.
 
## User-Level Aggregation
I then grouped the cleaned data by user, y_chromosome, and colorblind status to summarize each user's response accuracy.
Next, I calculated the average rank of the responses provided by each user.
Derived an accuracy score by I took  the inverse of the mean rank to derive accuracy score. This helps in ensuring  that lower ranks (more accurate naming) yielded higher scores.
For further filtering for reliability, I counted the total number of valid responses (total_answers) per user.

```{r data_cleaning, echo=FALSE}

user_accuracy_data <- answers %>%
  left_join(users, by = "user_id") %>%
  filter(!is.na(y_chromosome), !is.na(colorblind),spam_prob <= 0.5) %>%
  group_by(user_id, y_chromosome, colorblind) %>%
  summarise(
    mean_rank = mean(rank, na.rm = TRUE),
    accuracy_score = 1 / mean_rank, 
    user_total_responses = n(),
    .groups = "drop"
  ) 
describe(user_accuracy_data)
```
* Users answered a median of 5 questions and an average of 8.3. To ensure meaningful accuracy estimates, I removed users with fewer than 5 responses. This helps in reducing noise and improving the reliability of per-user accuracy scores. Having a sufficient number of observations is important

```{r further_data_cleaning, echo=FALSE}
user_accuracy_data<- user_accuracy_data %>%
  filter(user_total_responses >= 5)  
```




```{r reate-user-groups-sex-vision, echo=FALSE}
user_accuracy_data <- user_accuracy_data  %>%
  mutate(
    sex = ifelse(y_chromosome == 1, "Male", "Female"),
    vision = ifelse(colorblind == 1, "Colorblind", "Normal Vision"),
    user_group = paste(sex, vision, sep = ".")
  )

```

```{r anova-user-group-accuracy, echo=FALSE}
my_anova_model <- aov(accuracy_score ~ user_group, data = user_accuracy_data)
summary(anova_model)

```

```{r tukey-posthoc-user-group, echo=FALSE}
 TukeyHSD(my_anova_model)
```


```{r summarize-user-group-accuracy, echo=FALSE}
  group_means <- user_accuracy_data %>%
  group_by(user_group) %>%
  summarise(
    mean_accuracy = mean(accuracy_score),
    sd_accuracy = sd(accuracy_score),
    n = n(),
    se = sd_accuracy / sqrt(n),
    .groups = "drop"
  )
group_means
```
```{r t-test_for_sex_vision}

t_sex <- t.test(accuracy_score ~ sex, data = user_accuracy_data)
d_sex <- cohen.d(accuracy_score ~ sex, data = user_accuracy_data)

t_sex
d_sex

t_vision <- t.test(accuracy_score ~ vision, data = user_accuracy_data)
d_vision <- cohen.d(accuracy_score ~ vision, data = user_accuracy_data)
t_vision
d_vision
```


```{r compute-mean-lines-sex-vision, echo=FALSE}
mean_lines <- user_accuracy_data %>%
  group_by(sex, vision) %>%
  summarise(mean_acc = mean(accuracy_score, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    label = paste0("Mean: ", round(mean_acc, 3)),
    y_position = case_when(
      sex == "Female" ~ 7.5,
      sex == "Male" ~ 7
    )
  )


```


```{r animated-density-sex-by-vision, echo=FALSE}
animated_density <- ggplot(user_accuracy_data, aes(x = accuracy_score, fill = sex)) +
  geom_density(alpha = 0.5, aes(color = sex)) +  # apply both fill and color to sex
  geom_vline(data = mean_lines, 
             aes(xintercept = mean_acc, color = sex, group = sex),
             linetype = "dashed", size = 1) +
  geom_text(data = mean_lines, 
            aes(x = mean_acc, y = y_position, label = label, group = sex, color = sex),
            hjust = -0.1, size = 5) +  # remove show.legend = FALSE
  scale_fill_manual(values = c("Female" = "red", "Male" = "blue")) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(
    title = "Distribution of Accuracy Scores — {closest_state}",
    x = "Accuracy Score",
    y = "Density",
    fill = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    legend.position = "right"
  ) +
  transition_states(vision, transition_length = 2, state_length = 1) +
  ease_aes("cubic-in-out")
# View animated plot
 animate(animated_density, fps = 10, width = 800, height = 800)
```



```{r save_animation, echo=FALSE}
animated_gif <- animate(
  animated_density,
  fps = 10,
  width = 764,                  
  height = 400,
  renderer = gifski_renderer()
)
anim_save("accuracy_dist_animation.gif", animation = animated_gif)


```






