---
title: "tidytuesday_2025_08_05"
output: html_document
date: "2025-08-05"
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Income Inequality Before and After Taxes

This week we're exploring Income Inequality Before and After Taxes, as processed and visualized by Joe Hasell at Our World in Data: "Income inequality before and after taxes: how much do countries redistribute income?"

All data was processed by [Our World in Data]](https://ourworldindata.org), using these sources:

The Gini coefficient measures inequality on a scale from 0 to 1. Higher values indicate higher inequality. Inequality is measured here in terms of income before and after taxes and benefits.
Which countries have the highest Gini coefficient before taxes?
Which countries have the highest Gini coefficient after taxes?
Which countries have the highest shifts in Gini coefficient?
Which countries have the lowest shifts in Gini coefficient?
Which countries have had the highest changes in redistribution in the available data?
Thank you to Jon Harmon, Data Science Learning Community for curating this week's dataset.




# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load( tidyverse, tidytuesdayR, DataExplorer, tools, scales, janitor,openxlsx)

```


# Import the Data
```{r load_data, echo=FALSE}
# Import the #TidyTuesday data for 2025-08-05

tuesdata <- tidytuesdayR::tt_load('2025-08-05')
income_inequality_processed <- tuesdata$income_inequality_processed
income_inequality_raw <- tuesdata$income_inequality_raw

# Explore structure and completeness
introduce(income_inequality_processed)
introduce(income_inequality_raw)
plot_intro(income_inequality_processed)
plot_intro(income_inequality_raw)
glimpse(income_inequality_processed)
glimpse(income_inequality_raw)

# After initial exploration, I noticed that the datasets do not include key variables 
# such as Top 10% or Top 20% income shares, which are essential for the analysis I had in mind.
# Therefore, I decided to supplement the analysis by loading data directly from the OECD:
# https://www.oecd.org/content/dam/oecd/en/data/datasets/income-and-wealth-distribution-databases/idd-key-indicators.xlsx


```

```{r import_data_oecd, echo=FALSE}

# Import OECD Key Indicators on Income and Wealth Distribution
# The relevant data starts at row 6, and rows 6–44 include all available country-level observations.
# Earlier rows contain metadata, and later rows include notes or non-country entries I  don't need.
# I cleaned the column names for consistency and preview the structure.

oecd_df <- read.xlsx(
  "https://www.oecd.org/content/dam/oecd/en/data/datasets/income-and-wealth-distribution-databases/idd-key-indicators.xlsx", 
  sheet = 1, 
  startRow = 6,
  rows = 6:44
) %>%
  clean_names()

# Preview the structure of the data
glimpse(oecd_df)

```



# Data Preparation 
```{r data_prep, echo=FALSE}

# Select and rename only relevant indicators (e.g., Gini, S80/S20 ratio, top/bottom income shares, poverty rate)
names(oecd_df)[c(1, 2, 7, 12:18, 23)] <- c(
  "Entity", "Code", "gini_2022_or_latest", "S80_20_ratio",
  "Bottom_10_pct", "Bottom_20_pct", "Bottom_40_pct",
  "Top_40_pct", "Top_20_pct", "Top_10_pct", "Poverty_rate_2022"
)

# Subset the cleaned dataset to include only relevant columns
oecd_df_clean <- oecd_df[, c(1, 2, 7, 12:18, 23)]

# Preview structure before transformation
glimpse(oecd_df_clean)

# Convert character columns to numeric where needed and round for readability
oecd_df_clean <- oecd_df_clean %>%
  mutate(
    Entity = as.factor(Entity),
    Code = as.factor(Code),
    gini_2022_or_latest = round(as.numeric(gini_2022_or_latest), 3),
    across(
      .cols = c(S80_20_ratio, Bottom_10_pct, Bottom_20_pct, Bottom_40_pct,
                Top_40_pct, Top_20_pct, Top_10_pct, Poverty_rate_2022),
      ~ round(as.numeric(.x), 1)
    )
  )
```


```{r plot_data_prep, echo=FALSE}

# Prepare data for tornado plot: Top 20% vs Bottom 20% disposable income share
oecd_tornado <- oecd_df_clean %>%
  filter(!is.na(Top_20_pct), !is.na(Bottom_20_pct)) %>%
  select(Entity, Top_20_pct, Bottom_20_pct) %>%
  mutate(
    Top_20_pct = round(Top_20_pct, 1),
    Bottom_20_pct = round(Bottom_20_pct, 1)
  ) %>%
  pivot_longer(cols = c(Top_20_pct, Bottom_20_pct), names_to = "group", values_to = "value") %>%
  mutate(
    value = if_else(group == "Top_20_pct", value, -value),  # Flip Bottom 20% to left side
    group = recode(group, "Top_20_pct" = "Top 20%", "Bottom_20_pct" = "Bottom 20%")
  )

# Order countries by Top 20% share (ascending for plot)
entity_order <- oecd_tornado %>%
  filter(group == "Top 20%") %>%
  arrange(value) %>%
  pull(Entity)

oecd_tornado$Entity <- factor(oecd_tornado$Entity, levels = entity_order)

# Identify top 5 most unequal countries by Top 20% income share
top5_countries <- oecd_tornado %>%
  filter(group == "Top 20%") %>%
  arrange(desc(value)) %>%
  slice_head(n = 5) %>%
  pull(Entity)

# Build the tornado plot
plot1 <- ggplot(oecd_tornado, aes(x = value, y = Entity, fill = group)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("Top 20%" = "#1f77b4", "Bottom 20%" = "#ff7f0e")) +
  scale_x_continuous(
    limits = c(-100, 100),
    breaks = seq(-100, 100, by = 10),
    labels = abs,
    expand = expansion(mult = c(0, 0))
  ) +

  # Add data labels for Top 5 countries (Top 20% income share)
  geom_text(
    data = oecd_tornado %>% filter(Entity %in% top5_countries, group == "Top 20%"),
    aes(label = paste0(round(value, 1), "%")),
    hjust = -0.2,
    size = 3.5,
    color = "black"
  ) +

  # Annotation for Top 5 countries
  annotate("text",
           x = 65,
           y = match("Mexico", levels(oecd_tornado$Entity)),
           label = "These are five countries \nwhere the top 20% hold \nat least 45% of national \ndisposable income.",
           hjust = 0,
           size = 3.5,
           fontface = "italic"
  ) +

  # Plot labels and theme
  labs(
    title = "Disparities in Income Distribution: Top 20% vs Bottom 20% Shares in OECD Countries",
    subtitle = "Wider gaps = higher inequality: In Costa Rica, Chile, Türkiye, Mexico, and the U.S., the top 20% hold much \nmore than the bottom 20% (<7%). Narrower gaps in countries like Slovak Republic, Slovenia, and Belgium \nreflect more balanced income distribution.",
    caption = "Source: OECD Key Indicators on Income and Wealth Distribution",
    x = "Share of National Disposable Income (%)",
    y = NULL,
    fill = "Income Group"
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.y = element_text(face = "bold", size = 9),
    axis.title.x = element_text(size = 12, face = "bold", family = "Arial"),
    axis.text.x = element_text(size = 12, face = "bold", family = "Arial", hjust = 1),
    plot.caption = element_text(size = 10, hjust = 0),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  )

```








```{r build_plot, echo=FALSE}

# Build the tornado plot
plot1 <- ggplot(oecd_tornado, aes(x = value, y = Entity, fill = group)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("Top 20%" = "#1f77b4", "Bottom 20%" = "#ff7f0e")) +
  scale_x_continuous(
    limits = c(-100, 100),
    breaks = seq(-100, 100, by = 10),
    labels = abs,
    expand = expansion(mult = c(0, 0))
  ) +

  # Add data labels for Top 5 countries (Top 20% income share)
  geom_text(
    data = oecd_tornado %>% filter(Entity %in% top5_countries, group == "Top 20%"),
    aes(label = paste0(round(value, 1), "%")),
    hjust = -0.2,
    size = 3.5,
    color = "black"
  ) +

  # Annotation for Top 5 countries
  annotate("text",
           x = 65,
           y = match("Mexico", levels(oecd_tornado$Entity)),
           label = "These are five countries \nwhere the top 20% hold \nat least 45% of national \ndisposable income.",
           hjust = 0,
           size = 3.5,
           fontface = "italic"
  ) +
  labs(
    title = "Disparities in Income Distribution: Top 20% vs Bottom 20% Shares in OECD Countries",
    subtitle = "Wider gaps = higher inequality: In Costa Rica, Chile, Türkiye, Mexico, and the U.S., the top 20% hold much \nmore than the bottom 20% (<7%). Narrower gaps in countries like Slovak Republic, Slovenia, and Belgium \nreflect more balanced income distribution.",
    caption = "Source: OECD Key Indicators on Income and Wealth Distribution",
    x = "Share of National Disposable Income (%)",
    y = NULL,
    fill = "Income Group"
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.y = element_text(face = "bold", size = 9),
    axis.title.x = element_text(size = 12, face = "bold", family = "Arial"),
    axis.text.x = element_text(size = 12, face = "bold", family = "Arial", hjust = 1),
    plot.caption = element_text(size = 10, hjust = 0),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  )
```



```{r save_plot, echo=FALSE}
ggsave(
  plot = plot1,
  filename = "oecd_plot_linkedin.png",
  width = 12,    
  height = 6.3,  
  units = "in",
  dpi = 600
)

```


# Key Insights
- Wide income gaps persist in countries like Costa Rica, Chile, Türkiye, Mexico, and the U.S., where the top 20% hold over 45% of disposable income, and the bottom 20% hold less than 7%.
- Costa Rica stands out as the most unequal country in the dataset, with the top 20% holding over 50% of national disposable income.
- In contrast, countries such as Slovakia, Slovenia, Belgium, and Iceland show narrower gaps, with a more balanced distribution between top and bottom earners.
- Despite accounting for disposable income (after taxes and transfers), inequality remains pronounced in several OECD countries, suggesting limited redistribution impact in those contexts.
- Tornado plots are effective for revealing relative income share symmetry — highlighting both concentration at the top and marginalization at the bottom simultaneously.