---
title: "tidytuesday_2025_10_14"
output:
  html_document: default
  pdf_document: default
date: "2025-10-14"
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# FAO Food Security Indicators
This week’s #TidyTuesday dataset focuses on the Food and Agriculture Organization (FAO). FAO is celebrating 80 years of the world’s collective effort to end hunger.
The data comes from FAO’s Suite of Food Security Indicators, which tracks global patterns in food access, nutrition, and agricultural resilience.
I focused on one of the most urgent measures:
“Prevalence of severe food insecurity in the total population (percent, 3-year average)”
This indicator captures the share of people in each country living in households where adults have gone hungry — skipping meals, reducing portions, or going an entire day without eating due to lack of resources.

This analysis is both a celebration and a reminder:
Eighty years on, the world produces enough food — yet equitable access remains our greatest challenge.


# Load Required Libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(pacman)
p_load( tidyverse, tidytuesdayR, DataExplorer, tools, scales, janitor,ggbump,patchwork, slider) 

```

# Import the Data
```{r load_data, echo=FALSE}

tuesdata <- tidytuesdayR::tt_load('2025-10-14')

food_security <- tuesdata$food_security


# Explore structure and completeness
introduce (food_security)


glimpse(food_security )

head(food_security)

```



# Data Prep
```{r}

my_indicator <- "Prevalence of severe food insecurity in the total population (percent) (3-year average)"

# I wanted to keep only countries (excluding regions)
world_regions <- map_data("world") %>% distinct(region) %>% pull(region)

fs_countries <- food_security %>%
  transmute(
    year = as.integer(Year_End),
    area = Area,
    item = Item,
    value = as.numeric(Value)
  ) %>%
  mutate(region = case_when(
    area == "United States of America" ~ "USA",
    area == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
    area == "Russian Federation" ~ "Russia",
    area == "Côte d'Ivoire" ~ "Ivory Coast",
    area == "Iran (Islamic Republic of)" ~ "Iran",
    area == "Viet Nam" ~ "Vietnam",
    area == "Lao People's Democratic Republic" ~ "Laos",
    area == "Syrian Arab Republic" ~ "Syria",
    area == "Eswatini" ~ "Swaziland",
    area == "Türkiye" ~ "Turkey",
    area == "Republic of Moldova" ~ "Moldova",
    area == "Bolivia (Plurinational State of)" ~ "Bolivia",
    area == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
    area == "Democratic Republic of the Congo" ~ "Democratic Republic of the Congo",
    area == "China, mainland" ~ "China",
    TRUE ~ area
  )) %>%
  semi_join(tibble(region = world_regions), by = "region")

severe_fi <- fs_countries %>%
  filter(item == my_indicator, !is.na(value))

# the most recent year available (2024)
latest_year <- max(severe_fi$year, na.rm = TRUE)

# Prepare top 20 countries 
top20 <- severe_fi %>%
  filter(year == latest_year) %>%
  arrange(desc(value)) %>%
  slice_head(n = 20) %>%
  mutate(area = if_else(area == "Democratic Republic of the Congo", 
                        "DR Congo", 
                        as.character(area))) %>%
  mutate(area = fct_reorder(area, value))

```

# Plot 
```{r, plot, echo=FALSE}

# plot
ggplot(top20, aes(x = value, y = area)) +
  geom_col(fill = "#b22222", colour = "white", width = 0.7, linewidth = 0.2) +
  geom_text(aes(label = paste0(round(value, 1))), # , "%"
            hjust = -0.1, size = 3.5, color = "black") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = paste0("Eighty Years After 'Fiat Panis', Millions Still Go Hungry"),
    subtitle = "Top 20 Countries Facing Severe Food Insecurity",
    x = "Percent of Population",
    y = NULL,
    caption = "Source: FAO Suite of Food Security Indicators\n Note:  Values represent 3-year averages;  Includes only countries with data in 2024"
  ) +
theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7", colour = NA),
    plot.background  = element_rect(fill = "#f7f7f7", colour = NA),
    axis.text.y  = element_text(size = 10, face = "bold", family = "Arial",colour = "black"),
    axis.title.y = element_text(size = 10, face = "bold", family = "Arial"),
    axis.title.x = element_text(size = 10, face = "bold", family = "Arial"),
    strip.text   = element_text(face = "bold", size = 13),
    plot.title   = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle= element_text(size = 12, face = "bold", hjust = 0.5, lineheight = 1.2),
    axis.text.x  = element_text(size = 8, face = "bold", family = "Arial", hjust = 1), 
    plot.caption = element_text(size = 8,,hjust = 0, margin = margin(t = 8)), 
    plot.caption.position = "plot")
  




 ggsave("Top_20_Countries_Severe_Food_Insecurity.png", width = 9.5, height = 7.5, dpi = 320)
 
 

```


# Key Insights
- South Sudan leads with 63% of its population facing severe food insecurity — the world’s highest rate in 2024.
- Sub-Saharan Africa dominates the top ranks, with four of the top five countries located in the region.
- Eight decades after “Fiat Panis,” hunger remains widespread. Ending food insecurity is still unfinished work.